\documentclass[11pt]{article}
\usepackage[margin=0.95in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{mdframed}
\definecolor{accent}{HTML}{2563a8}
\definecolor{soft}{HTML}{e8eef5}
\newmdenv[linecolor=accent,linewidth=1pt,backgroundcolor=soft,roundcorner=4pt,
  innertopmargin=7pt,innerbottommargin=7pt,innerleftmargin=10pt,innerrightmargin=10pt]{keybox}
\newcommand{\R}{\mathbb{R}}
\newcommand{\vecop}{\operatorname{vec}}
\title{\textbf{OD Demand Compression as a Linear Encoder--Decoder}\\[2pt]
\large Folding, unfolding, and response-preserving zone aggregation}
\date{}
\begin{document}
\maketitle
\vspace{-3em}
\begin{center}\rule{0.5\textwidth}{0.4pt}\end{center}

Companion to \texttt{compress\_the\_response.tex} (the intuition) and
\texttt{superzone\_design\_principles.md} (the engineering). Here we make the
aggregation a precise linear system and use it to \emph{choose} the encoder.

\section{The assignment response chain}
OD demand $d\in\R^{n_{OD}}$, path flow $f\in\R^{n_P}$, link flow $v\in\R^{n_L}$:
\[
d \xrightarrow{R} f=Rd \xrightarrow{B^{\!\top}} v=B^{\!\top}f=\underbrace{B^{\!\top}R}_{H}\,d,
\qquad v = Hd .
\]
$R$ is the OD$\to$path split, $B$ the path--link incidence, and
$H=B^{\!\top}R\in\R^{n_L\times n_{OD}}$ is the \textbf{OD-to-link response}.
Link cost $\tau=c(v)$; linearised about equilibrium, $\Delta\tau\approx C_v\Delta v$
with $C_v=\operatorname{diag}(c_a'(v_a))$, so the OD-to-time response is $T=C_vH$.

\section{Encoder (fold) and decoder (unfold)}
Let $S\in\{0,1\}^{N\times K}$ map each original zone to one of $K$ super-zones
($S_{ik}=1$ iff zone $i\in$ super-zone $k$). Because an OD is a pair, the OD
folding matrix is the Kronecker square $Q=S\otimes S$, and with $d=\vecop(D)$,
\[
\textbf{encoder:}\quad E=Q^{\!\top}=(S\otimes S)^{\!\top},\qquad \bar d = Q^{\!\top}d
\quad(\text{super-zone OD}).
\]
A decoder $U$ unfolds the super-zone OD back to a loadable original OD,
$\tilde d = U\bar d = U Q^{\!\top} d$, giving a linear autoencoder
$P=UQ^{\!\top}$, $\ \tilde d = Pd$. Two decoders serve different goals:
\[
\textbf{reconstruction } U_R=Q(Q^{\!\top}Q)^{-1}\ (\text{best } d\approx U_RQ^{\!\top}d),
\qquad
\textbf{computational } U_C\ (\text{load super OD }(k,l)\text{ at representative }(c_k,c_l)).
\]
DTALite/TAPLite uses $U_C$. The \emph{hierarchical} build
(\texttt{superzone\_hier.py}) realises $U_C$ in the network itself: super-zones are
the only centroids, every original zone is a through node, zero-cost connectors
join \mbox{super $\leftrightarrow$ member}, so demand keyed to super-zones loads
through the members.

\section{The right error: response distortion, not OD reconstruction}
The quantity to preserve is the link flow, not the OD table:
\[
v=Hd,\qquad \tilde v = H\tilde d = HUQ^{\!\top}d,\qquad
\boxed{\,e_v = v-\tilde v = H\,(I-UQ^{\!\top})\,d\,}.
\]
``Lossless'' means $Hd\approx HUQ^{\!\top}d$, i.e. $H(I-UQ^{\!\top})d\approx0$ ---
\emph{not} $d\approx UQ^{\!\top}d$. Travel time follows,
$e_\tau\approx C_v\,e_v = C_vH(I-UQ^{\!\top})d$, bounded by
$\lVert e_\tau\rVert\le L_c\lVert e_v\rVert$ for locally Lipschitz $c$.

\begin{keybox}
\textbf{Corner-case theorem (verified).} If $S=I_N$ (one super-zone per zone) then
$Q=I$, the computational decoder $U_C$ is the identity loading, $UQ^{\!\top}=I$, so
$e_v=0$ \emph{exactly}. Empirically: Chicago Sketch and Chicago Regional at $S=I$
reproduce the full assignment to every decimal (link-volume $R^2=1.000$, $0\%$
RMSE, identical VMT/TT/gap). This is the unit test for any encoder/decoder.
\end{keybox}

\section{Low-rank structure and the corridor subspace}
$H$ is huge but low-rank: many OD pairs share corridors. With
$H\approx U_s\Sigma_sV_s^{\!\top}$ and the corridor projector $P_s=U_sU_s^{\!\top}$,
we need only the \emph{dominant} modes preserved,
\[
\boxed{\,P_sH(I-UQ^{\!\top})d\approx0\,}.
\]
\textit{Empirically (Chicago Regional, $N=1790$):} the demand matrix $D$ itself is
rank $\approx 32$ --- $\operatorname{SVD}(D)$ holds $50\%$ of the energy in rank $1$
(the gravity/size mode), $90\%$ in $3$, $99\%$ in $32$. So $K=178$ super-zones give
\emph{far} more capacity than the response needs: \textbf{the encoder rank is not the
binding constraint --- the decoder $U_C$ is} (intra-super demand it drops and the
single representative it loads through).

\section{Choosing the encoder $S$: experiment ($K{=}178$, 20 iters)}
Five constructions of $S$, each run and scored by response distortion vs.\ full:
\begin{center}\small
\begin{tabular}{lcccc}
\toprule
encoder $S$ & link $R^2$ & corridor (fwy) $R^2$ & \%RMSE & volume kept \\
\midrule
geographic quantile grid      & 0.808 & 0.892 & 73 & 0.690 \\
geographic $k$-means           & 0.802 & 0.951 & 76 & 0.656 \\
\textbf{demand-weighted $k$-means} & \textbf{0.849} & \textbf{0.966} & \textbf{66} & \textbf{0.708} \\
low-rank OD-SVD embedding      & 0.838 & 0.952 & 68 & 0.703 \\
flow-direction $k$-means       & 0.795 & 0.943 & 77 & 0.651 \\
\bottomrule
\end{tabular}
\end{center}
\noindent\textbf{Recommendation: demand-weighted $k$-means} (cluster zone centroids
with weights $=$ total zone demand). It wins every metric --- corridor $R^2=0.97$ ---
because the rank-1 mode \emph{is} the gravity/size pattern, and demand-weighting puts
super-zone resolution exactly where the response is largest. The
\textbf{low-rank OD-SVD embedding} (cluster origins by the top-$r$ left singular
vectors of $D$) is the principled runner-up and should overtake demand-weighting on
networks where corridors are \emph{not} aligned with demand density. Plain
geographic grids/$k$-means and direction signatures trail. All $k$-means variants
beat the quantile grid on corridors (more compact clusters).

\section{Design problem and the decoder lever}
The full statement:
\[
\boxed{\;\min_{S,\,U}\ \big\lVert P_sH(I-UQ^{\!\top})d\big\rVert_W^2
\;+\;\lambda\,I(S)\;+\;\mu\,C(S),
\qquad Q=S\otimes S,\ H=B^{\!\top}R,\ P_s=U_sU_s^{\!\top}\;}
\]
with $I(S)$ the intra-super-zone (dropped) demand and $C(S)$ a compactness/
interpretability penalty. Since the experiment shows the decoder dominates the
residual at usable $K$, the highest-value next move is a \textbf{better $U_C$}:
spread each super-zone's loading across its member access points in proportion to
member demand (a soft, demand-weighted $U_C$) instead of a single representative ---
this directly shrinks the $I(S)$ term and the single-access concentration error,
the two things the corner case removes.

\begin{keybox}
\textbf{Summary.} Zone aggregation is a linear autoencoder on the OD object:
encoder $Q^{\!\top}=(S\otimes S)^{\!\top}$ folds, decoder $U$ unfolds, and the only
error that matters is $e_v=H(I-UQ^{\!\top})d$ --- response, not reconstruction. The
OD response is rank $\sim\!32$, so capacity is abundant; pick $S$ by
\emph{demand-weighted} clustering (corridor $R^2\approx0.97$ at $K{=}178$), keep the
low-rank OD-SVD embedding as the response-aware alternative, and invest remaining
accuracy in a demand-spread decoder $U_C$.
\end{keybox}
\end{document}
